---
title: "Basic Coding Syntax and Structure in R"
author: "Norman Lo"
date: "5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.1 - Basic Mathematical Operations

As a programming language, R is built for basic mathematical operations.  We begin our exploration of R from these simple operations.  

|  Arithmetic Operators  |  Operations  |  Examples  |
|  :---:  |  :---:  |  :---:  |
|  +  |  Addition  |  15 + 5 = 20  |
|  -  |  Subtraction  |  15 - 5 = 10  |
|  *  |  Multiplication  |  15 * 5 = 75  |
|  /  |  Division  |  15 / 5 = 3  |
|  %/%  |  Integer Division  |  16%/% 3 = 5  |
|  ^  |  Exponent  |  15 ^ 3 = 3375  |
|  %%  |  Modulus  |  15 %% 5 = 0, 17%% 4 = 1  |

Here are some coding examples:

```{r Mathematical Operations 1}
# Addition
1 + 1 + 10

# Subtraction
10 - 5 - 1

# Multipication
3 * 2 * 4

# Division
10 / 5

# Exponent, returns the power of one variable against the other
2^4

# Modulus, returns the remainder after the division
17%%4

```

The above examples demonstrates the simple mathematical operations in R. It is worth to note that R is similar to other programming language, which follows the **order of operation**.  Here are some examples to demonstrate:

``` {r Mathematical Operations 2}
# Multiplication before addition
5 + 6 * 4

# Parenthese always goes first
(5 + 6) * 4

# Mixed with the parenthese with other operators
5 + 6 * (4 - 2)

```

Note: we have been using white space to separate the operators with the numbers in the previou examples, but it is not necessarily to leave a whitespace in coding.  However, it would be a good practice to improve the readability in your code.

## 1.2 - Variable Assignment



```{r pressure, echo=FALSE}
plot(pressure)
```




## 1.3 - Data Types






## 1.4 - Factor Vector




## 1.5 - R Basic Functions




## 1.6 - Missing Values




## 1.7 - Pipe









