---
title: "Reading Data into R"
author: "Norman Lo"
date: "6/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading Different Data Source into R

We have covered the basic features of R in the first two sections. In this section, we would like to explore how to read data from different data source into R for analysis. R offers several methods to read data from different sources, the most common one is reading from CSV (comma separated values) file and the basic excel file format. R also has the capability to read from relational database, internet data, and computer science data format "JSON". 

#### 3.1 - Reading CSV File

The best way to read a CSV file to R is using read.table() function. Many people starts with read.csv() function, but it's actually a wrapper function from read.table(), which is the same as using the **sep** arugment with a comma (",") in read.table() function. read.table() function returns a R data.frame object. 

The way to use read.table() function is to put the **absolute path** to the CSV file in the first argument. The path could be the location to the file in the machine or it could be a file from the internet. We are going to demonstrate to read a CSV file from the URL in this exercise.

``` {r CSV1}
# Assign the URL to a variable
theUrl <- "http://www.jaredlander.com/data/TomatoFirst.csv"

# Read the CSV file from the URL to R
tomato <- read.table(file=theUrl, header=TRUE, sep=",")

# Check the data.frame object
head(tomato)
```



``` {r CSV2}
# Create three individual vectors
x <- 10:1
y <- -4:5
q <- c("Hockey", "Football", "Baseball", "Curling", "Rugby",
       "Lacrosse", "Basketball", "Tennis", "Cricket", "Soccer")

# Create data.frame by using the three vectors
theDF <- data.frame(First=x, Second=y, Sport=q, stringsAsFactors=FALSE)

# Check the Sport column
theDF$Sport
```




``` {r CSV3}
# Loading readr package
library(readr)

# Assign URL to a variable
theUrl <- "http://www.jaredlander.come/data/TomatoFirst.csv"

# Loading the CSV to R and create a tibble object
tomato2 <- read_delim(file=theUrl, delim=", ")
tomato2
```



``` {r CSV4}
# Loading data.table package
library(data.table)

# Assign URL to a variable
theUrl <- "http://www.jaredlander.com/data/TomatoFirst.csv"

# Loading the CSV to R and create a data.table object
tomato3 <- fread(input=theUrl, sep=', ', header=TRUE)
head(tomato3)
```





#### 3.2 - Loading Excel Files


``` {r Excel1}
# Downlaoding an excel file from the internet
download.file(url='http://www.jaredlander.com/data/ExcelExample.xlsx',
              destfile='data/ExcelExample.xlsx', mode='wb')

# Loading readxl package
library(readxl)

# Loading the first sheet in the xlsx file from the data folder
excel_sheets('data/ExcelExample.xlsx')

# Create a tibble object and assign to a variable
tomatoXL <- read_excel('data/ExcelExample.xlsx')
tomatoXL
```



``` {r Excel2}
# Loading the second sheet in the xlse file from the data folder
wineXL1 <- read_excel('data/ExcelExample.xlsx', sheet=2)
head(wineXL1)

# Loading the second sheet in the xlse file by sheet name
wineXL2 <- read_excel('data/ExcelExample.xlsx', sheet='Wine')
head(winXL2)
```




#### 3.3 - Connecting to Database



``` {r Database1}
# Downloading database file
download.file("http://www.jaredlander.com/data/diamonds.db",
              destfile="data/diamonds", mode="wb")

# Loading RSQLite package
library(RSQLite)

# Creating a database driver
drv <- dbDriver('SQLite')
class(drv)

# Build connection to database
con <- dbConnect(drv, 'data/diamonds.db')
class(con)

# Extracting table names from the database
dbListTable(con)

# Extracting field (column) names from diamond table
dbListFields(con, name='diamonds')

# Extracting field (column) names from DiamondColors table
dbListFields(con, name='DiamondColors')
```








#### Reading Data from other Software Formats

``` {r Others}


```


#### Reading RData File

``` {r RData}

```


#### Loading the Built-In Data

``` {r Built-In Data}

```
