---
title: "Reading Data into R"
author: "Norman Lo"
date: "6/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading Different Data Source into R

We have covered the basic features of R in the first two sections. In this section, we would like to explore how to read data from different data source into R for analysis. R offers several methods to read data from different sources, the most common one is reading from CSV (comma separated values) file and the basic excel file format. R also has the capability to read from relational database, internet data, and computer science data format "JSON". 

#### 3.1 - Reading CSV File

The best way to read a CSV file to R is using read.table() function. Many people starts with read.csv() function, but it's actually a wrapper function from read.table(), which is the same as using the **sep** arugment with a comma (",") in read.table() function. read.table() function returns a R data.frame object. 

The way to use read.table() function is to put the **absolute path** to the CSV file in the first argument. The path could be the location to the file in the machine or it could be a file from the internet. We are going to demonstrate to read a CSV file from the URL in this exercise.

``` {r CSV1}
# Assign the URL to a variable
theUrl <- "http://www.jaredlander.com/data/TomatoFirst.csv"

# Read the CSV file from the URL to R
tomato <- read.table(file=theUrl, header=TRUE, sep=",")

# Check the data.frame object
head(tomato)
```

In this example, we use the three arguments file, header, and sep from the read.table() function to load the CSV file to R, which returns a data.frame object. Note that the argument names (file, header, and sep) are optional and can be missing in the function, such that read.table(theUrl, TRUE, ","), but it's a good habit to include the argument names for readability. The second argument "header=TRUE" reads the first row of the CSV file as the column names. The third argument 'sep=","' defines the data is separated by commas. We can also separate the data by different notations, such as "\t" (tab), ";" (semi-colon), etc, for different data formats.

In some cases, we may want to use "stringAsFactors=FALSE" argument to avoid a string column automatically read as factor. It saves the computational time when reading a large character data set. "stringAsFactor" can also be used in data.frame, here is an example to demonstrate.

``` {r CSV2}
# Create three individual vectors
x <- 10:1
y <- -4:5
q <- c("Hockey", "Football", "Baseball", "Curling", "Rugby",
       "Lacrosse", "Basketball", "Tennis", "Cricket", "Soccer")

# Create data.frame by using the three vectors
theDF <- data.frame(First=x, Second=y, Sport=q, stringsAsFactors=FALSE)

# Check the Sport column
theDF$Sport
```

When reading a large CSV file, read.table() function is relatively slow compares to other packages. For instance, we can use the read_delim() and data.table() from the **readr** package. The two functions are relatively quick and won't automatically transfrom character data to factor data type. 

``` {r CSV3}
# Loading readr package
library(readr)

# Assign URL to a variable
theUrl <- "http://www.jaredlander.come/data/TomatoFirst.csv"

# Loading the CSV to R and create a tibble object
tomato2 <- read_delim(file=theUrl, delim=", ")
tomato2
```

Similar to read.table(), the first argument of the read_delim() function is the path to the CSV file. The second argument col_names is default to be TRUE and the third argument delim=", " reads the data separated by commas. Note that read_delim() function or other data reading functions from readr package return a tibble object, which is a modern reimagining of the data.frame. In the readr package, read_csv, read_csv2, and read_tsv functions are the special cases, which separate the data by comma (, ), semi-colon (;), and tab (\t).

An alternative would be fread() function from data.table package. Similar to the previous data reading function, the first argument is the path or URL to the CSV file. The second argument sep=', ' represents the data separated by comma. The third argument header=TRUE reads the first row as the column names. This function also has the stringsAsFactor arugment, which is default to be FALSE.

``` {r CSV4}
# Loading data.table package
library(data.table)

# Assign URL to a variable
theUrl <- "http://www.jaredlander.com/data/TomatoFirst.csv"

# Loading the CSV to R and create a data.table object
tomato3 <- fread(input=theUrl, sep=', ', header=TRUE)
head(tomato3)
```

Genearlly speaking, both read_delim() and fread() are efficient functions for loading CSV file into R. The choice depends on the package (dplyr or data.table) that you are more fimilar to work with. 

#### 3.2 - Loading Excel Files


``` {r Excel1}
# Downlaoding an excel file from the internet
download.file(url='http://www.jaredlander.com/data/ExcelExample.xlsx',
              destfile='data/ExcelExample.xlsx', mode='wb')

# Loading readxl package
library(readxl)

# Loading the first sheet in the xlsx file from the data folder
excel_sheets('data/ExcelExample.xlsx')

# Create a tibble object and assign to a variable
tomatoXL <- read_excel('data/ExcelExample.xlsx')
tomatoXL
```



``` {r Excel2}
# Loading the second sheet in the xlse file from the data folder
wineXL1 <- read_excel('data/ExcelExample.xlsx', sheet=2)
head(wineXL1)

# Loading the second sheet in the xlse file by sheet name
wineXL2 <- read_excel('data/ExcelExample.xlsx', sheet='Wine')
head(winXL2)
```




#### 3.3 - Connecting to Database



``` {r Database1}
# Downloading database file
download.file("http://www.jaredlander.com/data/diamonds.db",
              destfile="data/diamonds", mode="wb")

# Loading RSQLite package
library(RSQLite)

# Creating a database driver
drv <- dbDriver('SQLite')
class(drv)

# Build connection to database
con <- dbConnect(drv, 'data/diamonds.db')
class(con)

# Extracting table names from the database
dbListTable(con)

# Extracting field (column) names from diamond table
dbListFields(con, name='diamonds')

# Extracting field (column) names from DiamondColors table
dbListFields(con, name='DiamondColors')
```








#### Reading Data from other Software Formats

``` {r Others}


```


#### Reading RData File

``` {r RData}

```


#### Loading the Built-In Data

``` {r Built-In Data}

```
